Supplementary file to "Varying coefficients correlated velocity models in complex landscapes with boundaries applied to narwhal responses to noise exposure"

### 2025-03-17 Latex/AoAS - Revision 2024. Alexandre, Adeline, Susanne/supp_revised.tex
add paper revision

Table~\ref{table: data distribution}  shows how the data is distributed among the different narwhals. Figure~\ref{fig: tracks before and after exposure} shows all the tracks before and after exposure with a base map of Scoresby Sound fjords system.

\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		Narwhal ID & Number of measurement before exposure & Number of measurements during exposure \\
		\hline
		A1 & 354 & 576\\
		\hline
		A2  & 151 & 515 \\
		\hline
		A3 & 397 & 680 \\
		\hline
		A4 & 127 & 642  \\
		\hline
		A5 & 207 & 419\\
		\hline
		A6 & 322 & 425 \\
		\hline
		\textbf{Total} & \textbf{1558} & \textbf{3257} \\
		\hline
	\end{tabular}
	\caption{Distribution of the data among the 6 individuals}
	\label{table: data distribution}
\end{table}

\begin{figure}[ht!]
	\centering
	\begin{subfigure}{0.8\linewidth}
		\centering
		\includegraphics[scale=0.5]{images/data_exploration/tracksBE2.png}
		\caption{Tracks before exposure experiments}
	\end{subfigure}

	\begin{subfigure}{0.8\linewidth}
		\centering
		\includegraphics[scale=0.5]{images/data_exploration/tracksAE.png}
		\caption{Tracks during exposure experiments}
	\end{subfigure}
	\caption{Movement data of East Greenland narwhals. The red crosses indicate initial positions.}
	\label{fig: tracks before and after exposure}
\end{figure}

All the relevant covariates   used for the analysis of narwhals movement are summarized in Table \ref{tab: summary covariates}.

\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|p{8cm}|c|}
		\hline
		Covariate & Unit & Description & Domain \\
		\hline
		$D^{ship}(t)$ & km & distance in kilometers between the narwhal and the ship at time $t$ & $\R^+$ \\
		\hline
		$E^{ship}(t)=\frac{1}{D^{ship}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
		\hline
		$D^{shore}(t)$ & km & distance between the narwhal and the nearest point on the shore at time $t$ & $\R^+$ \\
		\hline
		$I^{shore}(t)=\frac{1}{D^{shore}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the shore at time $t$ & $\R^+$ \\
		\hline
		$\Theta(t)$ & rad & angle between the vector that goes from the nearest shore point to the narwhal's position and the empirical velocity vector at time $t$  & $[-\pi,\pi]$ \\
		\hline 
	\end{tabular}
	\caption{Summary of the covariates}
	\label{tab: summary covariates}
\end{table}

@@

	It has Gaussian transition density with mean 
	\begin{equation}
		\E(V(t+\Delta \vert V(t)))=\exp(-A\Delta) V(t)+ (I_2-\exp(-A\Delta))\mu 
	\end{equation}
	and covariance matrix 
	\begin{align*}Var(V(t+\Delta) \vert V(t))&=\frac{4\nu^2}{\pi \tau} \int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) \exp(A(s-(t+\Delta)))^{\top} ds\\
		&=\frac{4\nu^2}{\pi \tau} \int_{0}^{\Delta} \exp(-Au) \exp(-Au)^{\top} du 
	\end{align*}
	Since $\exp(-Au)=\exp(-\frac{u}{\tau}) R_{-\omega u}$ where $R_{-\omega u}$ is the rotation matrix with angle $-\omega u$, the matrix product  $\exp(-Au) \exp(-Au)^{\top}$ is simply $\exp(-\frac{2u}{\tau}) I_2$. We deduce that the two components of the velocity are independent and have the same variance, denoted  $q_2(\Delta)$. The variance is 
	\begin{equation}
		q_2(\Delta)=\frac{2\nu^2}{\pi}\left(1-\exp\left(-\frac{2\Delta}{\tau}\right)\right).
	\end{equation}
	These results are found in \cite{gurarie_correlated_2017}.
	In the sequel, we  use the notation 
	\[\zeta(t,s) =\frac{2\nu}{\sqrt{\pi \tau}}\int_{t}^{s} \exp(A(u-s)) dW(u).\] 
	Using that $V(s)=\mu+\exp(-A(s-t))(V(t)-\mu)+\zeta(t,s)$, we have
\begin{align*}
    X(t+\Delta) &= X(t) + \int_t^{t+\Delta} V(s) \, ds \\
    &= X(t) + \mu \Delta + \int_t^{t+\Delta} \exp(-A(s-t))(X(t) - \mu) \, ds \\
    &\quad + \int_t^{t+\Delta} \zeta(t,s) \, ds \\
    % &= X(t) + \mu \Delta + \int_0^{\Delta} \exp(-As)(X(t) - \mu) \, ds + \int_t^{t+\Delta} \zeta(t,s) \, ds \\
    &= X(t) + \mu \Delta + \left(A^{-1}(V(t) - \mu) - A^{-1}\exp(-A\Delta)(X(t) - \mu)\right) \\
    &\quad + \int_t^{t+\Delta} \zeta(t,s) \, ds
\end{align*}
	Thus, 
	\begin{equation}
		X(t+\Delta)=X(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(V(t)-\mu)+\xi(t,t+\Delta) 
	\end{equation}
	where $ \xi(t,t+\Delta)=\int_t^{t+\Delta} \zeta(t,s)ds$.
	The location process is also Gaussian with mean 
	\begin{equation}
		\E(X(t+\Delta) \vert V(t),X(t)) =X(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(V(t)-\mu).
	\end{equation}
	To get an expression of the covariance matrix, first rewrite
	\begin{align*}
		\xi(t,t+\Delta)&=\int_t^{t+\Delta} \frac{2 \nu}{\sqrt{\pi \tau}} \left( \int_t^s \exp(-A((u-s))dW(u) \right) ds \\
		%&=\frac{2\nu}{\sqrt{\pi\tau}} \int_t^{t+\Delta} \left( \int_u^{t+\Delta} \exp(A(u-s)) ds \right) dW(u)  \\
		&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} (A^{-1}-A^{-1}\exp(A(u-t-\Delta))) dW(u) \\
		&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) dW(u)
	\end{align*}
	Then use Ito's isometry
 \begin{align*}
    \mbox{Var}(X(t+\Delta)\vert X(t),V(t))&=\frac{4 \nu^2}{\pi \tau} 
    \int_{t}^{t+\Delta} A^{-1} (I_2-\exp(A(u-t-\Delta)))\\
    &\quad \times (I_2-\exp(A(u-t-\Delta)))^\top A^{-\top} \, du \\
    &=\frac{4 \nu^2}{\pi \tau} \int_{0}^{\Delta} A^{-1} (I_2-\exp(-Ar))\\
    &\quad \times (I_2-\exp(-Ar))^\top A^{-\top} \, dr.
\end{align*}
This integral can be computed explicitly since 
\[A^{-1} (I_2-\exp(-Ar))(I_2-\exp(-Ar))^\top A^{-\top}=\frac{1}{C} f(r) I_2 \]
where $f(r)=1-2\exp\left( -\frac{r}{\tau}\right)\cos(\omega r)+\exp\left(\frac{-2r}{\tau}\right)$ and $C=\frac{1}{\tau^2}+\omega^2$.
We obtain that $X_1(t+\Delta)$ and $X_2(t+\Delta)$ are independent and have the same variance, denoted $q_1(\Delta)$.
Writing $\sigma=\frac{2\nu}{\sqrt{\pi \tau}}$, the variance is
\begin{align*}
    q_1(\Delta) &= \frac{\sigma^2}{C} \left( \Delta 
    - 2 \frac{\omega \sin(\omega \Delta) - \frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2} + \omega^2 } 
    \exp\left( -\frac{\Delta}{\tau} \right) \right. \\
    &\quad \left. + \frac{\tau}{2} \left( \frac{\omega^2 - \frac{3}{\tau^2}}{\frac{1}{\tau^2} + \omega^2} 
    - \exp\left( -\frac{2\Delta}{\tau}\right) \right) 
    \right)
\end{align*}

Now we compute the covariance between   $X$ and $V$ to get the full covariance matrix of $U$:
\begin{align*}
    \Gamma(\Delta) &= \frac{4\nu^2}{\pi \tau} \E\left(\left( \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) \, dW(u)\right) \right. \\
    &\quad \left. \times \left(\int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) \, dW(s)\right)^\top\right) \\
    &= \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-(t+\Delta)))) \exp(A(u-(t+\Delta)))^\top \, du \\
    &= \frac{4\nu^2}{\pi \tau} \int_0^{\Delta} A^{-1}(I_2-\exp(-Ar)) \exp(-Ar)^\top \, dr.
\end{align*}
	Then, 
	\[A^{-1}(I_2-\exp(-Ar)) \exp(-Ar)^\top=\frac{1}{C}\exp\left(-\frac{r}{\tau} \right) \begin{pmatrix} g(r) & h(r) \\ -h(r) & g(r) \end{pmatrix}\]
	where 
\begin{align*}
    g(r) &= \frac{1}{\tau} \left(\cos(\omega r) + \exp\left( -\frac{r}{\tau} \right)\right) 
    - \omega \sin(\omega r), \\
    h(r) &= -\frac{1}{\tau} \sin(\omega r) 
    + \omega \left(\cos(\omega r) - \exp\left( -\frac{r}{\tau}\right)\right).
\end{align*}
	Finally we get
	\[
	\gamma_1(\Delta)=\frac{\sigma^2}{2C}\left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega \Delta)\right),
	\]
	\[\gamma_2(\Delta)=\frac{\sigma^2}{C}\left( \exp\left( -\frac{\Delta}{\tau}\right) \sin(\omega \Delta)-\frac{\omega \tau}{2} \left(1-\exp\left( -2 \frac{\Delta}{\tau}\right) \right)\right).\]

@@

>>>>

Table~\ref{table: data distribution}  shows how the data is distributed among the different narwhals.
\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		Narwhal ID & Number of measurement before exposure & Number of measurements during exposure \\
		\hline
		A1 & 354 & 576\\
		\hline
		A2  & 151 & 515 \\
		\hline
		A3 & 397 & 680 \\
		\hline
		A4 & 127 & 642  \\
		\hline
		A5 & 207 & 419\\
		\hline
		A6 & 322 & 425 \\
		\hline
		\textbf{Total} & \textbf{1558} & \textbf{3257} \\
		\hline
	\end{tabular}
	\caption{Distribution of the data among the 6 individuals}
	\label{table: data distribution}
\end{table}

The relevant covariates   used for the analysis of narwhals movement are summarized in Table \ref{tab: summary covariates}.

\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|p{8cm}|c|}
		\hline
		Covariate & Unit & Description & Domain \\
		\hline
		$D^{ship}(t)$ & km & distance in kilometers between the narwhal and the ship at time $t$ & $\R^+$ \\
		\hline
		$E^{ship}(t)=\frac{1}{D^{ship}(t)}$ & $km^{-1}$ & global exposure level of the narwhal to the ship disturbance at time $t$ & $\R^+$ \\
		\hline
		$D^{shore}(t)$ & km & distance between the narwhal and the nearest point on the shore at time $t$ & $\R^+$ \\
		\hline
		$\Theta(t)$ & rad & angle between the vector that goes from the nearest shore point to the narwhal's position and the empirical velocity vector at time $t$  & $[-\pi,\pi]$ \\
		\hline 
	\end{tabular}
	\caption{Summary of the covariates}
	\label{tab: summary covariates}
\end{table}

@@

	It has Gaussian transition density with mean 
	\begin{equation}
		\E(V(t+\Delta \vert V(t)))=\exp(-A\Delta) V(t)+ (I_2-\exp(-A\Delta))\mu 
	\end{equation}
	and covariance matrix 
	\begin{align*}Var(V(t+\Delta) \vert V(t))&=\frac{4\nu^2}{\pi \tau} \int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) \exp(A(s-(t+\Delta)))^{\top} ds\\
		&=\frac{4\nu^2}{\pi \tau} \int_{0}^{\Delta} \exp(-Au) \exp(-Au)^{\top} du 
	\end{align*}
	An integration by part gives 
	\[\int_{0}^{\Delta} \exp(-Au) \exp(-Au)^{\top} du =A^{-1}(I_2-\exp(-A\Delta))(I_2-\exp(-A\Delta))^\top A^{-\top}\]. In the sequel, we denote $S(\Delta)=A^{-1}(I_2-\exp(-A\Delta))$.
	Since $\exp(-Au)=\exp(-\frac{u}{\tau}) R_{-\omega u}$ where $R_{-\omega u}$ is the rotation matrix with angle $-\omega u$, we can deduce that the two components of the velocity are independent and have the same variance, denoted  $q_2(\Delta)$. This variance is 
	\begin{equation}
		q_2(\Delta)=\frac{2\nu^2}{\pi}\left(1-\exp\left(-\frac{2\Delta}{\tau}\right)\right).
	\end{equation}
	These results are found in \cite{gurarie_correlated_2017}.
	In the sequel, we  use the notation 
	\[\zeta(t,s) =\frac{2\nu}{\sqrt{\pi \tau}}\int_{t}^{s} \exp(A(u-s)) dW(u).\] 
	Using that $V(s)=\mu+\exp(-A(s-t))(V(t)-\mu)+\zeta(t,s)$, we have
\begin{align*}
    X(t+\Delta) &= X(t) + \int_t^{t+\Delta} V(s) \, ds \\
    &= X(t) + \mu \Delta + \int_t^{t+\Delta} \exp(-A(s-t))(V(t) - \mu) \, ds \\
    &\quad + \int_t^{t+\Delta} \zeta(t,s) \, ds \\
    % &= X(t) + \mu \Delta + \int_0^{\Delta} \exp(-As)(X(t) - \mu) \, ds + \int_t^{t+\Delta} \zeta(t,s) \, ds \\
    &= X(t) + \mu \Delta + \left(A^{-1}(V(t) - \mu) - A^{-1}\exp(-A\Delta)(X(t) - \mu)\right) \\
    &\quad + \int_t^{t+\Delta} \zeta(t,s) \, ds
\end{align*}
	Thus, 
	\begin{equation}
		X(t+\Delta)=X(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(V(t)-\mu)+\xi(t,t+\Delta) 
	\end{equation}
	where $ \xi(t,t+\Delta)=\int_t^{t+\Delta} \zeta(t,s)ds$.
	The location process is also Gaussian with mean 
	\begin{equation}
		\E(X(t+\Delta) \vert V(t),X(t)) =X(t)+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(V(t)-\mu).
	\end{equation}
	To get an expression of the covariance matrix, first rewrite
	\begin{align*}
		\xi(t,t+\Delta)&=\int_t^{t+\Delta} \frac{2 \nu}{\sqrt{\pi \tau}} \left( \int_t^s \exp(-A((u-s))dW(u) \right) ds \\
		%&=\frac{2\nu}{\sqrt{\pi\tau}} \int_t^{t+\Delta} \left( \int_u^{t+\Delta} \exp(A(u-s)) ds \right) dW(u)  \\
		&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} (A^{-1}-A^{-1}\exp(A(u-t-\Delta))) dW(u) \\
		&= \frac{2\nu}{\sqrt{\pi \tau}} \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) dW(u)
	\end{align*}
	Then use Ito's isometry
 \begin{align*}
    \mbox{Var}(X(t+\Delta)\vert X(t),V(t))&=\frac{4 \nu^2}{\pi \tau} 
    \int_{t}^{t+\Delta} A^{-1} (I_2-\exp(A(u-t-\Delta)))\\
    &\quad \times (I_2-\exp(A(u-t-\Delta)))^\top A^{-\top} \, du \\
    &=\frac{4 \nu^2}{\pi \tau} \int_{0}^{\Delta} A^{-1} (I_2-\exp(-Ar))\\
    &\quad \times (I_2-\exp(-Ar))^\top A^{-\top} \, dr.
\end{align*}
This integral can be computed explicitly since 
\[A^{-1} (I_2-\exp(-Ar))(I_2-\exp(-Ar))^\top A^{-\top}=\frac{1}{C} f(r) I_2 \]
where $f(r)=1-2\exp\left( -\frac{r}{\tau}\right)\cos(\omega r)+\exp\left(\frac{-2r}{\tau}\right)$ and $C=\frac{1}{\tau^2}+\omega^2$.
We obtain that $X_1(t+\Delta)$ and $X_2(t+\Delta)$ are independent and have the same variance, denoted $q_1(\Delta)$.
Writing $\sigma=\frac{2\nu}{\sqrt{\pi \tau}}$, the variance is
\begin{align*}
    q_1(\Delta) &= \frac{\sigma^2}{C} \left( \Delta 
    - 2 \frac{\omega \sin(\omega \Delta) - \frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2} + \omega^2 } 
    \exp\left( -\frac{\Delta}{\tau} \right) \right. \\
    &\quad \left. + \frac{\tau}{2} \left( \frac{\omega^2 - \frac{3}{\tau^2}}{\frac{1}{\tau^2} + \omega^2} 
    - \exp\left( -\frac{2\Delta}{\tau}\right) \right) 
    \right)
\end{align*}

Now we compute the covariance between   $X$ and $V$ to get the full covariance matrix of $U$:
\begin{align*}
    \Gamma(\Delta) &= \frac{4\nu^2}{\pi \tau} \E\left(\left( \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-t-\Delta))) \, dW(u)\right) \right. \\
    &\quad \left. \times \left(\int_{t}^{t+\Delta} \exp(A(s-(t+\Delta))) \, dW(s)\right)^\top\right) \\
    &= \int_t^{t+\Delta} A^{-1}(I_2-\exp(A(u-(t+\Delta)))) \exp(A(u-(t+\Delta)))^\top \, du \\
    &= \frac{4\nu^2}{\pi \tau} \int_0^{\Delta} A^{-1}(I_2-\exp(-Ar)) \exp(-Ar)^\top \, dr.
\end{align*}
	Then, 
	\[A^{-1}(I_2-\exp(-Ar)) \exp(-Ar)^\top=\frac{1}{C}\exp\left(-\frac{r}{\tau} \right) \begin{pmatrix} g(r) & h(r) \\ -h(r) & g(r) \end{pmatrix}\]
	where 
\begin{align*}
    g(r) &= \frac{1}{\tau} \left(\cos(\omega r) - \exp\left( -\frac{r}{\tau} \right)\right) 
    + \omega \sin(\omega r), \\
    h(r) &= -\frac{1}{\tau} \sin(\omega r) 
    + \omega \left(\cos(\omega r) - \exp\left( -\frac{r}{\tau}\right)\right).
\end{align*}
	Finally we get
	\[
	\gamma_1(\Delta)=\frac{\sigma^2}{2C}\left( 1+\exp\left( -\frac{2\Delta}{\tau}\right)-2\exp\left( -\frac{\Delta}{\tau}\right) \cos(\omega \Delta)\right),
	\]
	\[\gamma_2(\Delta)=\frac{\sigma^2}{C}\left( \exp\left( -\frac{\Delta}{\tau}\right) \sin(\omega \Delta)-\frac{\omega \tau}{2} \left(1-\exp\left( -2 \frac{\Delta}{\tau}\right) \right)\right).\]

@@

\section{Approximate transition density}

We now suppose that the angular velocity $\omega$ varies with time. Denote $\mathcal{A}(0,t)=\int_0^t A(s) ds$ the time integral of the drift matrix.
A similar calculation as before gives
\[V(t+\Delta)=\exp(-\mathcal{A}(t,t+\Delta)) V(t)+\sigma \int_t^{t+\Delta} \exp(-\mathcal{A}(s,t+\Delta)) dW(s)\]

Therefore, $V(t+\Delta) \vert V(t)$ is normally distributed with mean
\[\E(V(t+\Delta) \vert V(t))=\exp(-\mathcal{A}(t,t+\Delta)) V(t)\]
and covariance matrix 
\[\mbox{Var}(V(t+\Delta) \vert V(t))=\sigma^2\int_t^{t+\Delta} \exp(-\mathcal{A}(s,t+\Delta)) \exp(-\mathcal{A}(s,t+\Delta))^\top ds\]

The exact formula for the process $X$ is 
\[X(t+\Delta)=X(t)+\int_t^{t+\Delta} \exp(-\mathcal{A}(t,s)) V(t) ds +\sigma \int_t^{t+\Delta} \int_t^s \exp(-\mathcal{A}(u,s)) dW(u) ds \]
Consequently, the process is gaussian with mean given by
\[\E(X(t+\Delta) \vert X(t))=X(t)+\int_t^{t+\Delta} \exp(-\mathcal{A}(t,s)) V(t) ds\]
and covariance matrix 
\[\mbox{Var}(X(t+\Delta) \vert X(t))=\sigma^2 \int_t^{t+\Delta} \lbrace \int_u^{t+\Delta} \exp(-\mathcal{A}(u,s))ds \rbrace \lbrace \int_u^{t+\Delta} \exp(-\mathcal{A}(u,s)) ds \rbrace^\top du\]

Denote $U(t)=(X(t), V(t))$. It is a Markov process with transition density
\[U(t+\Delta) \vert U(t)=u \sim \mathcal{N}(T(\Delta) u, Q(\Delta))\]
where 
\begin{equation}
	T(\Delta) = \begin{pmatrix}
		I_2 & \int_t^{t+\Delta} \exp(-\mathcal{A}(t,s)) ds \\
		0_2 & \exp(-\mathcal{A}(t,t+\Delta))
	\end{pmatrix}
\end{equation}

\begin{equation}
	Q(\Delta) = \begin{pmatrix}

		Q_{11}(\Delta) & Q_{12}(\Delta)  \\
		Q_{21}(\Delta) &  Q_{22}(\Delta)
	\end{pmatrix}
\end{equation}
where 
\[Q_{11}(\Delta)=\sigma^2 \int_{t}^{t+\Delta} \left( \int_{u}^{t+\Delta} \exp(-\mathcal{A}(u,s)) ds \right) \left( \int_{u}^{t+\Delta} \exp(-\mathcal{A}(u,s)) ds \right)^\top du \]
\[Q_12(\Delta)=\sigma^2 \int_{t}^{t+\Delta} \left( \int_{u}^{t+\Delta} \exp(-\mathcal{A}(u,s)) ds \right) \exp(-\mathcal{A}(u,t+\Delta))^\top du\]
\[Q_{21}(\Delta)=Q_{12}(\Delta)^\top\]
\[Q_{22}(\Delta)=\sigma^2 \int_t^{t+\Delta} \exp(-\mathcal{A}(s,t+\Delta)) \exp(-\mathcal{A}(s,t+\Delta))^\top ds\]

Suppose we have approximations of the time varying matrix $A$ at times $t_0=t < t_1 < \cdots < t_n=t+\Delta$ and denote $\Delta_i=t_{i+1}-t_i$ for $i \in \{0,\cdots,n-1\}$.
Then we can approximate
\[\exp(-\mathcal{A}(t,t+\Delta))=\prod_{i=0}^{n-1} \exp(-A_i\Delta_i)\]
where $A_i=A(t_i)$. Each matrix $\exp(-A_i\Delta_i)$ is a weighted rotation matrix with angle $-\omega_i\Delta_i$.
For conciseness, in the sequel we denote $E_i=\exp(-A_i\Delta_i)$, $P_{jk}= \prod_{i=j}^{k-1} E_i$ and $S_j=S(\Delta_j)$.

The covariance matrix of $V$ is approximated by 
\[Q_{22}(\Delta)\simeq\sum_{j=0}^{n-1} \sigma^2 P_{j+1,n} P_{j+1,n}^\top S_j S_j^\top\]
The mean for the process $X$ can be estimated with the following formula
\[\E(X(t+\Delta) \vert X(t)) \simeq X(t)+\sum_{k=0}^{n-1} P_{0,k} S_k  V(t) \]
The covariance can also be approximated. The first step is to approximate $\mathcal{A}(u,s)$ for $u \in [t,t+\Delta]$ and $s \in [u,t+\Delta]$. For a fixed $u \in [t_{j-1},t_j]$,

\begin{align*}
	& \exp(-\mathcal{A}(u,s)) \simeq 
	\exp(-A_{j-1}(t_j-u)) \1_{[u,t_j]}(s) +\\
	& \sum_{k=j}^{n-1} \exp(-A_{j-1}(t_j-u)) \times \prod_{i=j}^{k-1} \exp(-A_i \Delta_i) \times \exp(-A_k (s-t_k)) \1_{[t_k,t_{k+1}]}(s)
\end{align*}

This gives the following approximation for any $u \in [t_{j-1},t_j]$
\begin{align*}
	&\int_u ^{t+\Delta} \exp(-\mathcal{A}(u,s)) ds \simeq A_{j-1}^{-1} (I_2-\exp(-A_{j-1} (t_j-u)))+ \\
	&\sum_{k=j}^{n-1} \exp(-A_{j-1} (t_j-u)) \prod_{i=j}^{k-1} \exp(-A_i \Delta_i) A_k^{-1} (I_2- \exp(A_k \Delta_k)) 
\end{align*}
We then deduce an approximation of the covariance matrix.

\begin{align*}
	\mbox{Var}(X(t+\Delta) \vert X(t)) \simeq &\sum_{j=1}^n \sigma^2 A_{j-1}^{-1} A_{j-1}^{-\top}(\Delta_{j-1} I_2-S_{j-1}-S_{j-1}^\top + S_{j-1}S_{j-1}^\top) \\
	&+ \sigma^2  M_{jn}^\top A_{j-1}^{-1} (S_{j-1}^\top -S_{j-1} S_{j-1}^\top)M_j M_j^\top + \sigma^2  M_{jn} A_{j-1}^{-\top} (S_{j-1} -S_{j-1} S_{j-1}^\top) \\
	&+M_{jn} M_{jn}^\top S_{j-1} S_{j-1}^\top 
\end{align*}

where $M_{jn} = \sum_{k=j}^{n-1} P_{jk} S_k $.

Finally, we have the following formula to approximate the covariance matrix between $X$ and $V$.

\begin{align*}
	\mbox{Var}(X(t+\Delta),V(t+\Delta) \vert X(t),V(t))=& \sum_{j=1}^n \sigma^2 P_{j,n}^\top A_{j-1}^{-1}(S_{j-1}^\top-S_{j-1} S_{j-1}^\top)  
	+\sigma^2 M_{jn} P_{j,n}^\top S_{j-1} S_{j-1}^\top 
\end{align*}

We denote
\[\tilde{T}_n=\begin{pmatrix} I_2 & \sum_{k=0}^{n-1} P_{0,k} S_k \\
	0_2  &  \prod_{i=0}^{n-1}  E_i \end{pmatrix}
\]
and \[\tilde{Q}_n = \begin{pmatrix}  \tilde{Q}_n^{11}& \tilde{Q}_n^{12} \\
	\tilde{Q}_n^{21}& \tilde{Q}_n^{22} \end{pmatrix}\]
where 
\begin{align*}
	\tilde{Q}_n ^{11}=&\sum_{j=1}^n \sigma^2 A_{j-1}^{-1} A_{j-1}^{-\top}(\Delta_{j-1} I_2-S_{j-1}-S_{j-1}^\top + S_{j-1}S_{j-1}^\top) \\
	&+ \sigma^2  M_{jn}^\top A_{j-1}^{-1} (S_{j-1}^\top -S_{j-1} S_{j-1}^\top) + \sigma^2  M_{jn} A_{j-1}^{-\top} (S_{j-1} -S_{j-1} S_{j-1}^\top) \\
	&+\sigma^2M_{jn} M_{jn}^\top S_{j-1} S_{j-1}^\top \\
	\tilde{Q}_n ^{12}=&\sum_{j=1}^n \sigma^2 P_{j,n}^\top A_{j-1}^{-1}(S_{j-1}^\top-S_{j-1} S_{j-1}^\top)  
	+\sigma^2 M_{jn} P_{j,n}^\top S_{j-1} S_{j-1}^\top \\
	\tilde{Q}_n ^{21}=&(\tilde{Q}_n ^{12})^\top \\
	\tilde{Q}_n^{22} = & \sum_{j=0}^{n-1} \sigma^2 P_{j+1,n} P_{j+1,n}^\top S_j S_j^\top
\end{align*}
We can express $\tilde{T}_n$ and $\tilde{Q}_n$ in terms of the elementary matrices 
\[T_i=\begin{pmatrix} I_2 & S_i \\ 0_2 & E_i \end{pmatrix}  \mbox {and } 
Q_i=\begin{pmatrix} \sigma^2 A_i^{-1}A_i^{-\top} ( \Delta_iI_2-S_i-S_i^\top+S_iS_i^\top) & \sigma^2 A_i^{-1} (S_i^\top-S_iS_i^\top) \\
	\sigma^2 A_i^{-\top} (S_i-S_iS_i^\top) & \sigma^2 S_i S_i^\top
\end{pmatrix}
\]

\begin{equation}
	\tilde{T}_n = T_{n-1} \times \cdots \times T_1 \times T_0
	\label{eq: approximate link matrix}
\end{equation}
\begin{equation}
	\tilde{Q}_n= \sum_{k=1}^{n-1} T_{n-1}^\top \cdots T_{k+1}^\top Q_k T_{k+1} \cdots T_{n-1}
\end{equation}

\begin{proof}
	$\tilde{T}_n=T_0$ and $\tilde{Q}_1=Q_0$. Then, for $n \geq 2$, 
	\begin{align*}
		\tilde{T}_{n}=&\begin{pmatrix} 
			I_2 & P_{0,n-1}S_{n-1}+\sum_{j=0}^{n-2} P_{0,j} S_j \\
			0_2 & P_{0,n} E_{n-1}
		\end{pmatrix}
	=\begin{pmatrix}
		I_2 & S_{n-1} \\
		0_2 & E_{n-1}
	\end{pmatrix}
	\begin{pmatrix} 
	I_2 & \sum_{j=0}^{n-2} P_{0,j} S_j \\
	0_2 & P_{0,n-1}
	\end{pmatrix}\\
	&= T_{n-1} \tilde{T}_{n-1}
	\end{align*}
Then, using that $M_{n,n}= 0_2 $ and $P_{n,n}=I_2$ we get
\[ \tilde{Q}_n=Z_{n-1}+Q_{n-1} \]
where 
\[Z_{n-1}=\begin{pmatrix} \sum_{j=1}^{n-1} Q_{j-1}^{11} +M_{jn}^\top Q_{j-1}^{12}+M_{jn} Q_{j-1}^{21}+M_{j,n}M_{j,n}^\top Q_{j-1}^{22} & \sum_{j=1}^{n-1}  P_{j,n}^\top Q_{j-1}^{12} +M_{j,n} P_{j,n}^\top Q_{j-1}^{22} \\
	\sum_{j=1}^{n-1} P_{j,n} Q_{j-1} ^{21} + M_{jn}^\top P_{jn} Q_{j-1}^{22} & \sum_{j=1}^{n-1} P_{jn}^\top P_{jn} Q_{j-1}^{22}
	\end{pmatrix}
\]
Then, since $M_{j,n}=M_{j,n-1}+P_{j,n-1} S_{n-1}$ we get
\begin{align*} Z_{n-1}=\begin{pmatrix} \tilde{Q}_{n-1}^{11}+S_{n-1}^\top Q_{n-1}^{12}+S_{n-1} \tilde{Q}_{n-1}^{21}+S_{n-1}S_{n-1}^\top \tilde{Q}_{n-1}^{22} &  E_{n-1}^\top \tilde{Q}_{n-1}^{12}+E_{n-1}^\top S_{n-1} \tilde{Q}_{n-1}^{12}\\
		E_{n-1} \tilde{Q}_{n-1}^{21}+E_{n-1} S_{n-1}^\top \tilde{Q}_{n-1}^{22} & E_{n-1}E_{n-1}^\top
	\end{pmatrix}
\end{align*}
And this is equal to $T_{n-1}^\top \tilde{Q}_{n-1} T_{n-1}$.
